Automated inference of production rules for glycans

### 2021-07-06 papers/sugar-synth/algo.tex
fig spacing
15 page fit

In this section, we present a method to solve the synthesis problem of finding
a set of production rules that produce a given set $\mu$ of molecules.
Our synthesis method \textsc{SugarSynth} is presented in Algorithm~\ref{alg:sugar-synth}.
Here, we are considering only the single compartment case.
The generalizations are discussed in
\choosefinal{the extended version~\cite{arxiv-this}.}{ Appendix~\ref{sec:variations}.}

@@

We use templates to model the search space of rules.
A template is also a tree that has a depth and
the internal nodes of the tree have the same number of children.
Two variables label each node of the template.
One variable is for choosing the sugar at the node and the other is for describing
the `situation' of the node.
% The first variable at the node encodes choices of
% sugar monomers or their absence.
The domain of the first variables is $S \union \{\bot\}$.
Let $SVars$ be the unbounded set of variables with the domain.
We will use the pool of $SVars$ to add variables to the templates.

\begin{wrapfigure}{r}{0.35\textwidth}

        \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid,scale=0.5]
    \node[loc] (v0) {$v_0$};
    \node[sqloc, below right of=v0] (v1) {$v_e$};
    \draw [->] plot [smooth] coordinates { (v0.south)
          ($(v0.south)+(3mm,-2mm)$)
          ($(v0.south)+(10mm,1mm)$)
          % ($(v0.south)+(9mm,-6mm)$)
          % ($(v0.east)+(8mm,-8mm)$)
          (v1.north) };
    % \path[->] (v0) edge (v1);
    % \path[->,dashed] (v1) edge (v2);
    % \path[->] (v1) edge (v3);
    \fill[fill=gray,opacity=0.2] 
    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
    \fill[fill=gray,opacity=0.4] 
    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
  \end{tikzpicture}
  \caption{Parts of production rules in the rule templates.}
  \label{fig:rule-parts}
  \vspace{-8mm}
\end{wrapfigure}

A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations.
The first situation is when a node position is in the expanding part.
The dark gray nodes are the expanding nodes.
The second situation is when a node position is not in the rule.
The dashed area are the absent nodes. Among the matching nodes, we have two cases.
The third situation is when a node position is in the matching part
and has expanding descendants.
The nodes on the solid path from $v_0$ to the root $v_e$ of the expanding part are in the third situation.
% the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is the rest of the node positions
in the matching part, which is the light gray area.
% The nodes in the rest of the light gray area are in the fourth situation.
A variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the set of four situations.
Let $KVars$ be the unbounded set of variables with domain $K$. 
% We will use the pool of $KVars$ to add variables to the templates.
Our templates are sufficiently expressive to cover all aspects of biology.
The templates are defined as follows.

@@

At line~\ref{line:consNewR}, we construct constraints $rCons$ again using
$\textsc{EnocdeProduce}$ that says that template molecule $\hat{m}$
is generated by rules $R$.
We check the satisfiability of $rCons \land mCons$.
If it is not satisfiable, we have found the rules that generate exactly
the molecules in $\mu$ and the loop terminates.
Otherwise, we use the satisfying assignment to create a
counterexample molecule $m'$.
At line~\ref{line:negCons}, we add constraints to $nCons$ stating that
all possible applications of template rules in $T$ must not produce $m'$.
We use shorthand $F \landplus G$ for $ F := F \land G$.
As we will see that $\textsc{EncodeProduce}$ introduces fresh variable
maps $\tau$ and $cuts$ in the encoding.
Since we negate the returned formula by $\textsc{EncodeProduce}$ 
and then we check the satisfiability.
% the encoding will not be correct
% due to the fresh variables.
Therefore, we need to introduce universal quantifiers for the
fresh variables.
We introduce universal quantifiers over $\tau$ and $cuts$
variables before adding to $nCons$.
% Intuitively, the universal quantifiers say that all possible
% application patterns of the learned rules do not produce $m'$.
Afterwards, the loop goes to the next iteration.
Since the domain of all the variables is finite, eventually the loop must
terminate.

@@

In $\textsc{EncodeP}$, we can traverse up from $v$ for $\ell$ steps to
find the node to match the root of $t$.
Naturally, it returns $\lfalse$ if there
is no $\ell$th ancestor of $v$. % i.e., there is no match.
The variable $mark$ is the timestamp for node $v$.
% During matching, we will add constraints stating that the matching part of the
% template should match to the nodes added before $mark$
% and expanding part of the template should match with nodes that
% are added at or after $mark$.
The local variable $c$ collects the conjunction of constraints as they are
generated. % and $c$ is initially $\ltrue$.
The local variable $v_r$ is initially equal to the root $v_{0r}$ of $t$
and traverses the nodes in $t$.
The while loop at line~\ref{line:encodep-while} in $\textsc{EncodeP}$ starts with
the $\ell$th ancestor from $v$,
matches all the ancestors up to the parent of $v$.
As the loop traverses down, it also traverses $t$ along the matching
using variable $v_r$.
In each iteration, $v_r$ is updated to the $j$th child due to lines
\ref{line:encodep-next-vr} and \ref{line:encodep-update-vr}
if the ancestors of $v$ also traverse along with the $j$th child.
% (see if condition at line \ref{line:encodep-jth-child-cond}).

\vspace{-3mm}
\input{alg-encode-produce.tex}

% Let us consider the body of the loop.
Let $v'$ be the $i$th ancestor at some iteration of the loop.
At line~\ref{line:encodep-ans-match}, the loop adds constraints stating that
the corresponding node $v_r$ in the template rule is
$MatchAns$, sugar matches in $v_r$ and $v'$, and
node $v'$ is added before $mark$.
The loop at line~\ref{line:encodep-sub-for-loop}, iterates over
children of $v'$.
If there is a child of $v'$ that is not along the path to
$v$, it is matched at line~\ref{line:mtree} with the corresponding
child of $v_r$ by calling $\textsc{MatchTree}$, which traverses
rule template and molecule and generate constraints to encode that their
nodes match.
We will discuss $\textsc{MatchTree}$ shortly.
% $\textsc{MatchTree}$ is also called with a third parameter
% $mark$ and a Boolean value as the fourth parameter.
% We will discuss the procedure shortly.
If the $j$th child of $v'$ is along the path to $v$,
we get node $v'_r$ for updating
$v_r$ for the next iteration of the while loop.
% After the while loop in $\textsc{EncodeP}$, $v_r$ is the
% candidate node for the root of expanding nodes.
After the while loop in $\textsc{EncodeP}$ at line \ref{line:encodep-vr-expand}, we declare $v_r$
is $Expand$,  match the node $v$
with the corresponding node $v_r$ in the template rule by calling
$\textsc{MatchTree}$, and also call to match the cut pattern at
the subtree of $v$ with the template rule.

@@

In the case when we call \textsc{EncodeProduce} with template molecule and
concrete rules.
We need to swap the roles of $\nu$ and $M$ at their occurrences
at line \ref{line:encodep-ans-match} in \textsc{EncodeP} and
line \ref{line:matchtree-cons} in \textsc{MatchTree}.
Furthermore, $\kappa(v_r)$ is a concrete value depending on the situation of $v_r$ in
the rule.
In the functions, the variable is to be replaced by the
evaluated value of $\kappa(v_r)$.

\begin{theorem}
  If \textsc{SugarSynth}($\mu$, $d$, $n$) returns a set of rules $R$, then if $R$ produces a molecule that has depth less than $h+1$, the it is in $\mu$ (soundness).
  If it fails to synthesize rules, there are no rule set from the budget of depth $d$ and number $n$ (completeness).
\end{theorem}

The soundness is true due to the construction of constraints.
Our algorithm always terminates. At each iteration, at least one solution, i.e,
a set of rules is discarded.
In fact, we discard many because in each iteration we reject production of a counterexample molecule, but this may discard whole range of rule sets. 
Since the set of all possible rules is finite, we
guarantee termination.
Therefore, the completeness holds.
In the worst case, the entire search space is explored.

>>>>

In this section, we present a method to solve the synthesis problem of finding
a set of production rules that produce a given set $\mu$ of molecules.
Our method \textsc{SugarSynth} is Algorithm~\ref{alg:sugar-synth}.
Here, we are considering only the single compartment case.
The generalizations are discussed in
\choosefinal{the extended version~\cite{arxiv-this}.}{ Appendix~\ref{sec:variations}.}

@@

We use templates to model the search space of rules.
A template is a tree that has a depth and
the internal nodes of the tree have the same number of children.
Two variables label each node. %of the template.
One variable is for choosing the sugar at the node and the other is for describing
the `situation' of the node.
% The first variable at the node encodes choices of
% sugar monomers or their absence.
The domain of the first variables is $S \union \{\bot\}$.
Let $SVars$ be the unbounded set of variables with the domain.
We will use the pool of $SVars$ to add variables to the templates.

A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations.
The first situation is when a node is in the expanding part, which are shown in 
dark gray. % nodes are the expanding nodes.
The second situation is when a node is not in the rule.
The dashed area are the absent nodes.
Among the matching nodes, we have two cases.
\begin{wrapfigure}{r}{0.35\textwidth}
  % \vspace{-12mm}
        \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid,scale=0.5]
    \node[loc] (v0) {$v_0$};
    \node[sqloc, below right of=v0] (v1) {$v_e$};
    \draw [->] plot [smooth] coordinates { (v0.south)
          ($(v0.south)+(3mm,-2mm)$)
          ($(v0.south)+(10mm,1mm)$)
          % ($(v0.south)+(9mm,-6mm)$)
          % ($(v0.east)+(8mm,-8mm)$)
          (v1.north) };
    % \path[->] (v0) edge (v1);
    % \path[->,dashed] (v1) edge (v2);
    % \path[->] (v1) edge (v3);
    \fill[fill=gray,opacity=0.2] 
    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
    \fill[fill=gray,opacity=0.4] 
    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
  \end{tikzpicture}
  \caption{Parts of production rules in the rule templates.}
  \label{fig:rule-parts}
  \vspace{-8mm}
\end{wrapfigure}
The third situation is when a node is in the matching part
and has expanding descendants.
The nodes on the solid path from $v_0$ to the root $v_e$ of the expanding part are in the third situation.
% the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is the rest of the nodes
in the matching part, which is in light gray.
% The nodes in the rest of the light gray area are in the fourth situation.
A variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the situations.
Let $KVars$ be an unbounded set of variables with domain $K$. 
% We will use the pool of $KVars$ to add variables to the templates.
Our templates are sufficiently expressive to cover all aspects of biology, which are defined as follows.

@@

At line~\ref{line:consNewR}, we construct constraints $rCons$ again using
$\textsc{EnocdeProduce}$ that says that template molecule $\hat{m}$
is generated by rules $R$.
We check the satisfiability of $rCons \land mCons$.
If it is not satisfiable, we have found the rules that generate exactly
the molecules in $\mu$ and the loop terminates.
Otherwise, we use the satisfying assignment to create a
counterexample molecule $m'$.
At line~\ref{line:negCons}, we add constraints to $nCons$ stating that
all possible applications of template rules in $T$ must not produce $m'$.
We use shorthand $F \landplus G$ for $ F := F \land G$.
As we will see that $\textsc{EncodeProduce}$ introduces fresh variable
maps $\tau$ and $cuts$ in the encoding.
Since we negate the returned formula by $\textsc{EncodeProduce}$ 
and then we check the satisfiability,
% the encoding will not be correct
% due to the fresh variables.
we need to introduce universal quantifiers over the
fresh variables. %before adding to $nCons$.
% Intuitively, the universal quantifiers say that all possible
% application patterns of the learned rules do not produce $m'$.
Afterwards, the loop goes to the next iteration.
% Since the domain of all the variables is finite, eventually the loop must
% terminate.

@@

In $\textsc{EncodeP}$, we can traverse up from $v$ for $\ell$ steps to
find the node to match the root of $t$.
% Naturally,
It returns $\lfalse$ if there
is no $\ell$th ancestor of $v$. % i.e., there is no match.
The variable $mark$ is the timestamp for $v$.
% During matching, we will add constraints stating that the matching part of the
% template should match to the nodes added before $mark$
% and expanding part of the template should match with nodes that
% are added at or after $mark$.
The variable $c$ collects the  constraints as they are
generated. % and $c$ is initially $\ltrue$.
The variable $v_r$ is initially equal to the root $v_{0r}$ of $t$
and traverses the nodes in $t$.
The while loop at line~\ref{line:encodep-while} starts with
the $\ell$th ancestor from $v$,
matches all the ancestors up to the parent of $v$.
As the loop traverses down, it also traverses $t$ 
using variable $v_r$.
In each iteration, $v_r$ is updated to the $j$th child due to lines
\ref{line:encodep-next-vr} and \ref{line:encodep-update-vr}
if the ancestors of $v$ also traverse along the $j$th child.
% (see if condition at line \ref{line:encodep-jth-child-cond}).

\vspace{-10mm}
\input{alg-encode-produce.tex}

% Let us consider the body of the loop.
Let $v'$ be the $i$th ancestor at some iteration of the loop.
At line~\ref{line:encodep-ans-match}, the loop adds constraints stating that
the corresponding node $v_r$ in $t$ is
$MatchAns$, sugar matches in $v_r$ and $v'$, and
node $v'$ was added before $mark$.
The loop at line~\ref{line:encodep-sub-for-loop}, iterates over
children of $v'$.
If there is a child of $v'$ that is not along the path to
$v$, it is matched at line~\ref{line:mtree} with the corresponding
child of $v_r$ by calling $\textsc{MatchTree}$, which
will be discussed shortly.
% traverses
% rule template and molecule and generate constraints to encode that their
% nodes match.
% We will discuss $\textsc{MatchTree}$ shortly.
% $\textsc{MatchTree}$ is also called with a third parameter
% $mark$ and a Boolean value as the fourth parameter.
% We will discuss the procedure shortly.
If the $j$th child of $v'$ is along the path to $v$,
we get node $v'_r$ for updating
$v_r$ for the next iteration of the while loop.
% After the while loop in $\textsc{EncodeP}$, $v_r$ is the
% candidate node for the root of expanding nodes.
After the while loop in $\textsc{EncodeP}$ at line \ref{line:encodep-vr-expand}, we declare $v_r$
is $Expand$,  match the node $v$
with the corresponding node $v_r$ in the template rule by calling
$\textsc{MatchTree}$, and also call to match the cut pattern at
the subtree of $v$ with the template rule.

@@

In the case when we call \textsc{EncodeProduce} with template molecule and
concrete rules.
We swap the roles of $\nu$ and $M$ at their occurrences
at line \ref{line:encodep-ans-match} in \textsc{EncodeP} and
line \ref{line:matchtree-cons} in \textsc{MatchTree}.
Furthermore, $\kappa(v_r)$ is a concrete value.
% depending on the situation of $v_r$ in the rule.
In the functions, the variable is to be replaced by the
evaluated value of $\kappa(v_r)$.

\begin{theorem}
  If \textsc{SugarSynth}($\mu$, $d$, $n$) returns  rules $R$, then if $R$ produces a molecule that has depth less than $h+1$, the it is in $\mu$ (soundness).
  If it fails to find rules, there are no rule sets within the budget of $d$ and  $n$ (completeness).
\end{theorem}

The soundness is true by construction.
% Our algorithm always terminates.
At each iteration, at least one solution, i.e,
a set of rules is discarded.
In fact, we discard many because in each iteration we reject all rule sets that may produce a counterexample molecule. 
Since the set of all possible rules is finite, we terminate.
Therefore, the completeness holds.
% In the worst case, the entire search space is explored.

### 2021-07-06 papers/sugar-synth/experiments.tex
dsadsa
15 page fit

In this section, we present our implementation of the proposed method.
% %
% We also apply our tool on several data sets and illustrate the usability of our method.

{\em Implementation:} 
We have implemented method \textsc{SugarSynth} in a prototype tool {\ourtool}.
%
The tool, written in {\tt C++}, uses {\zthree}\cite{z3} as the SMT solver
to discharge the generated satisfiability queries.
%
% We have developed a specialized file format for giving input to~\ourtool.
% %
% The format allows one to declare monomers, list molecules over the monomers
% in the SMTLIB like format,
% and pass the input parameters of the method.

@@

{\em Results:}
We have applied \ourtool~on the data set. For each data set, we choose several
parameter combinations to illustrate the relative performance of the tool.
If we did not budget large enough parameters such as the size of unknown molecules, number of rules etc, then the tool fails to
synthesize the rules.
We present the rules learned after giving minimum resources.
% Our tool gave output that matched with the analysis of~\cite{Jaiman2018}
% in reasonable time.

For D1, we synthesize the rules in 1.60 seconds. Even by reducing the first two parameters, we were able to synthesize the rules but it took longer time.
Giving an extra compartment in the third row did not impact the performance.  We also observe the trade-off between the number of compartments
and the depth of the rules at the second and third row of D1 and how it impacts performance.
We synthesize the rules for D2, involving runaway reactions,  in 7.97 seconds.
For D2, we also learned large rules, i.e., they are adding many nodes at a single time.
We synthesize the rules for D3 in 0.57 seconds. We synthesize the rules for D4, which is also our motivating example, in 0.85 seconds.
However, if we use too many resources or too little, the tool runs for a long time as the search in combinatorial space is highly sensitive to the parameters.
The synthesis for D5 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
Our synthesized production is in line with the reported rules
in the literature~\cite{Jaiman2018}.
More experiments on the variations of the problem
are included in the appendix~\ref{sec:ex-variants}. %\hyperref[sec:appendix]{}.

>>>>

% In this section, we present our implementation of the proposed method.
% %
% We also apply our tool on several data sets and illustrate the usability of our method.

{\em Implementation:} 
We have implemented \textsc{SugarSynth} in a tool {\ourtool}~\cite{Charles2013}.
%
The tool, written in {\tt C++}, uses {\zthree}\cite{z3} as the SMT solver
to solve the satisfiability queries. We ran the experiments  on a laptop with 8GB RAM and 2.4GHz CPU.
%
% We have developed a specialized file format for giving input to~\ourtool.
% %
% The format allows one to declare monomers, list molecules over the monomers
% in the SMTLIB like format,
% and pass the input parameters of the method.

@@

{\em Results:}
We have applied \ourtool~on the data set. For each data set, we choose several
parameter combinations to illustrate the relative performance of the tool.
If we did not budget large enough parameters such as the size of unknown molecules, number of rules etc, then the tool fails to
synthesize the rules.
We present the rules learned after giving minimum resources. However, the set of rules reported need not be either unique or minimal.
% Our tool gave output that matched with the analysis of~\cite{Jaiman2018}
% in reasonable time.

For D1, we synthesize the rules in 1.60 seconds. Even by reducing the first two parameters, we were able to synthesize the rules but it took longer time.
Giving an extra compartment in the third row did not impact the performance.  We also observe the trade-off between the number of compartments
and the depth of the rules at the second and third row of D1 and how it impacts performance.
We synthesize the rules for D2, involving runaway reactions,  in 7.97 seconds.
For D2, we also learned large rules, i.e., they are adding many nodes at a single time.
We synthesize the rules for D3 in 0.57 seconds. We synthesize the rules for D4, which is also our motivating example, in 0.85 seconds.
However, if we use too many resources or too little, the tool runs for a long time as the search in combinatorial space is highly sensitive to the parameters.
The synthesis for D5 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
Our synthesized production is in line with the reported rules
in the literature~\cite{Jaiman2018}.
\choosefinal{}{
More experiments on the variations of the problem
are included in the Appendix~\ref{sec:ex-variants}.} %\hyperref[sec:appendix]{}.

### 2021-07-06 papers/sugar-synth/model.tex
minor changes
15 page fit

Let $R$ be a set of rules.
A molecule $m$ is {\em producible} by $R$ from a set of molecules $Q$
if there is sequence of molecules $m_0,...,m_k$
such that $m_0 \in Q$, $m_k=m$, and
for each $0<i\leq k$, $m_{i} = Apply(m_{i-1},r)$ for some $r \in R$.
Let $P(Q,R)$ denote the set of molecules that are producible from
rules $R$ from a set of molecules $Q$.
We have discussed in section~\ref{sec:bio} that all the production rules
are not applied at the same time.
The rules may live in compartments and
the rule sets of the compartments are applied one after another.
To model compartments for the rules,
let us suppose we have a sequence $R_1,...,R_k$ of set of rules.
Let $P(Q, R_1,..,R_k) = P(..P(P(Q,R_1),R_2),..,R_k)$ denoting the
trees obtained after applying the rule sets one after another.

% \subsection{Synthesis problem}
In nature, we observe a set of glycan molecules $\mu$ present in a cell.
%
However, we do not necessarily know the production rules to produce the molecules.
%
We will be developing a method to find the production rules.
%
The {\em synthesis problem} is to find a set $R$ of production rules
such that $\mu = P(S,R)$,
where $S$ is the set of monomers.

>>>>

Let $R$ be a set of rules.
A molecule $m$ is {\em producible} by $R$ from a set of molecules $Q$
if there is sequence of molecules $m_0,...,m_k$
such that $m_0 \in Q$, $m_k=m$, and
for each $0<i\leq k$, $m_{i} = Apply(m_{i-1},r)$ for some $r \in R$.
Let $P(Q,R)$ denote the set of molecules that are producible from
rules $R$ from a set of molecules $Q$.
We have discussed in Section~\ref{sec:bio} that all the production rules
are not applied at the same time.
The rules may live in compartments and
the rule sets of the compartments are applied one after another.
To model compartments for the rules,
let us suppose we have a sequence $R_1,...,R_k$ of set of rules.
Let $P(Q, R_1,..,R_k) = P(..P(P(Q,R_1),R_2),..,R_k)$ denoting the
trees obtained after applying the rule sets one after another.

% \subsection{Synthesis problem}
In nature, we observe a set of glycan molecules $\mu$ present in a cell.
%
However, we may not know the production rules to produce the molecules.
%
We will be developing a method to find the rules.
%
The {\em synthesis problem} is to find a set $R$ of production rules
such that $\mu = P(S,R)$,
where $S$ is the set of monomers.
